---
title: "Activity Monitoring Data Assessment"
author: "Lisa Seghers"
date: "September 6, 2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "figs/fig-")
```


### Loading and preprocessing the data

First, the dataset is read in from the working directory.  Any relevant libraries are also loaded.

``` {r}
library(readr)
library(dplyr)
library(ggplot2)
activity <- read_csv("activity.csv")
```

Then the dataset is evaluated to see if preprocessing is necessary prior to initial analysis.  
It's a straightforward dataset with 3 variables with clear names and 17568 rows.  
Although the dataset contains NAs, these can be ignored for the first few steps.



### What is the mean total number of steps taken per day?

Find the total number of steps taken for each day.

``` {r steptotals}
dailysteps <- aggregate(steps ~ date, activity, sum, na.rm = TRUE)
```

Make a histogram of the total number of steps taken per day:

``` {r hist1}
par(mar = c(5, 5, 2, 1))
hist(dailysteps$steps, 
     main = "Histogram of Total Steps per Day", 
     xlab = "Step Totals")
```

Calculate and report the mean and median of the total number of steps taken per day.

``` {r initialstats}
meanSteps <- mean(dailysteps$steps)
medianSteps <- median(dailysteps$steps)
meanStepsNum <- format(meanSteps, scientific = FALSE)
medianStepsNum <- format(medianSteps, scientific = FALSE)
```

The mean number of steps taken per day is `r meanStepsNum`.  
The median number of steps taken per day is `r medianStepsNum`.



### What is the average daily activity pattern?

Find average number of steps per interval, collapsed across all days.

``` {r stepintervals}
intervalSteps <- aggregate(cbind(steps, date) ~ interval, activity, mean, na.rm = TRUE)
```

Show the time series plot as a line graph:

``` {r intervalplot}
with(intervalSteps, plot(interval, steps, type = "l", lwd = 2,
                         xlab = "5 Minute Intervals", 
                         ylab = "Average Steps",
                         main = "Average Steps per 5 Minute Interval Across All Days"))
```

``` {r maxinterval}
maxInterval <- which(intervalSteps$steps == max(intervalSteps$steps))
maxInt <- intervalSteps$interval[maxInterval]
```

The 5-minute interval with the highest average number of steps is `r maxInt`.



### Imputing missing values

Determine the total number of missing values in the dataset.

``` {r missing}
missing <- sum(is.na(activity$steps))
```

There are `r missing` NAs in the steps column.

The selected strategy to impute the missing values is to replace NAs with the mean for that interval.

``` {r impute}
activity$meanSteps <- intervalSteps$steps[match(activity$interval, intervalSteps$interval)]
activity$steps1 <- for (i in 1:17568) {
      if(is.na(activity$steps[i])) { 
        activity$steps[i] <- activity$meanSteps[i]
      }
}
```

Calculate the new daily step totals with imputed valued included.

Make a histogram of the updated total number of steps taken each day.

``` {r hist2}
dailystepsImputed <- aggregate(cbind(steps, interval, meanSteps) ~ date, 
                               activity, sum, na.rm = TRUE)
par(mar = c(5, 5, 2, 1))
hist(dailystepsImputed$steps, 
     main = "Histogram of Total Steps per Day (with imputed values)", 
     xlab = "Step Totals")
```

Calculate and report the mean and median of the total number of steps taken per day, including imputed values.

``` {r imputedstats}
meanStepsImputed <- mean(dailystepsImputed$steps)
medianStepsImputed <- median(dailystepsImputed$steps)
meanStepsImpNum <- format(meanStepsImputed, scientific = FALSE)
medianStepsImpNum <- format(medianStepsImputed, scientific = FALSE)
```

The mean number of steps taken per day, including imputed values, is `r meanStepsImpNum`.  
The median number of steps taken per day, including imputed values, is `r medianStepsImpNum`.

These numbers are basically identical to the previous numbers due to the use of the interval means to replace the NAs.  The imputed values biased the data toward the mean/median (the center bar of the histogram gets noticeably taller while the other bars get a little shorter).



### Are there differences in activity patterns between weekdays and weekends?

Add a new factor variable to the dataset (with imputed valued) - two levels, "weekday" and "weekend".  


``` {r weekfactor}
fullActivity <- activity[, 1:3]
fullActivity$date <- weekdays(fullActivity$date)
weekend <- c("Saturday", "Sunday")
fullActivity$week <- ifelse(fullActivity$date %in% weekend, "weekend", "weekday")
```

Find the means of steps by interval, grouped by weekend and weekday.

``` {r meanbyday}
intervalSteps2 <- aggregate(steps ~ interval + week, fullActivity, mean)
```

Show the line graphs of mean steps per interval for weekday and weekend as a two row panel graph:

``` {r weekplot}
g <- ggplot(intervalSteps2, aes(interval, steps))
p <- g + geom_line(aes(color = week), size = 1.2) + facet_wrap(week~., nrow = 2) + theme_bw()
print(p)
```

Weekdays have an earlier rise in activity around 5:30 am and a higher spike of activity between 8 and 9am than weekends, but weekends show more consistent activity all day during waking hours.



